---
title: "Lab 2 Assignment: Group 02"
format:
  html:
    embed-resources: true
editor: visual
---

## Load libraries 

We load the libraries necessary for the PCA analysis

```{r}


library(tidyverse)
library(broom)  
library(cowplot)
```

## Load data

We load the data from breast cancer biopsies.

```{r}

biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")


```

## PCA analysis

We make a PCA analysis, by using the function prcomp:

```{r}

pca_fit <- biopsy %>% 
  select(where(is.numeric)) |>  
  prcomp(scale = TRUE)
```

To show how the data is projected on our two most principal components, we make a scatter plot.

```{r}
pca_fit %>%
  augment(biopsy) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#CC79A7", benign = "#56B4E9")
  ) +
  theme_half_open(12) + background_grid()+
  labs(
    title = "PCA of Biopsy Data",
    subtitle = "PC1 vs PC2 colored by outcome",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Diagnosis Outcome" 
  )
```

Now we want to see each data groups on the PCA coordinates

```{r}

pca_fit %>%
  tidy(matrix = "rotation")
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style, color = "#56B4E9") +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  labs(
    title = "PCA Rotation Matrix",
    subtitle = "PC1 vs PC2 with variable loadings",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
```

Now we want to see how much variance is explained by each principal component. This serves to see how much variance is explained by each attribute.

```{r}

pca_fit %>%
  tidy(matrix = "eigenvalues")

pca_fit %>%
  tidy(matrix = "eigenvalues") %>%
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#D55E00", alpha = 0.8) + 
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    title = "Variance Explained by Principal Components",
    subtitle = "Percentage of variance for each PC",
    x = "Principal Components",
    y = "Variance Explained (%)"
  ) +
  theme_minimal_hgrid(12)
```
